---
title: "How Americans Like Their Steak"
author: "John Cruz"
date: "2023-01-27"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE)
```

## Overview

Walt Hickey from FiveThirtyEight collected data from people within the United States to see if a risk-averse person would be more likely to order a steak well done. They found no evidence a person that was a higher risk taker would prefer their steaks rare. 

['FiveThirtyEight Article'](https://fivethirtyeight.com/features/how-americans-like-their-steak/)

['Data Source'](https://raw.githubusercontent.com/fivethirtyeight/data/master/steak-survey/steak-risk-survey.csv)

## Required libraries
```{r library, message=FALSE}
library(pander)
library(tidyverse)
library(pollster)
```

## Data Dictionary

The original data set columns included the full questions asked. They will be renamed with keywords for ease:

```{r data_dictionary-table, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
# install.packages('pander')

tabl <- "
| Keyword  | Original Column   |
|:--------|:-----------------------------|
| id      | RespondentID |
| lottery_pick | Consider the following hypothetical situations: <br> &emsp;&emsp;In **Lottery A**, you have a 50% chance of success, with a payout of $100. <br> &emsp;&emsp;In **Lottery B**, you have a 90% chance of success, with a payout of $20. <br> Assuming you have $10 to bet, would you play Lottery A or Lottery B?      |
| smoker | Do you ever smoke cigarettes?      |
| alcohol | Do you ever drink alcohol?      |
| gamble | Do you ever gamble?     |
| skydiving | Have you ever been skydiving?      |
| speeding | Do you ever drive above the speed limit?
| cheated | Have you ever cheated on your significant other?      |
| eat_steak | Do you eat steak?      |
| doneness | How do you like your steak prepared?      |
| gender | Gender      |
| age_group | Age      |
| income_level | Household Income      |
| education_level | Education      |
| us_region | Location (Census Region)      |
"
pander(tabl) # output the table in a format good for HTML/PDF/docx conversion
```

## Data Preparation
Import data from GitHub and rename columns

```{r import-data-rename-cols}
new_colnames <- c('id', 'lottery_pick', 'smoker', 'alcohol', 'gamble', 'skydiving', 'speeding', 'cheated', 'eat_steak', 'doneness', 'gender', 'age_group', 'income_level', 'education_level', 'us_region')

steak_survey <- read.csv('https://raw.githubusercontent.com/fivethirtyeight/data/master/steak-survey/steak-risk-survey.csv', col.names = new_colnames)
```

View first few rows within the data frame.

```{r top-rows}
knitr::kable(head(steak_survey, 5))
```

Remove data that is explicitly NA from the *id* column and people who do not eat steak. Subset data to focus on risks and their region. 

```{r remove-na-rows, results='hide'}
steak_survey <- steak_survey %>% 
  drop_na(id)
```

```{r subset-cols}
subset_cols <- c('smoker', 'alcohol', 'gamble', 'skydiving', 'speeding', 'cheated', 'doneness', 'us_region')
risks <- steak_survey %>% 
  select(all_of(subset_cols))
```

```{r remove-any-null-cols, warning=FALSE}
non_null_risks <- risks %>% 
  filter(across(all_of(subset_cols), ~ .x != ''))

knitr::kable(head(non_null_risks, 5))
```

## Findings
We can see that most individuals prefer Medium and Medium Rare cooking temperatures for their steaks. 

```{r barplot}
ggplot(data = non_null_risks) +
  geom_bar(aes(doneness))
```

Calculate total risks taken by each individual.

```{r groupby-doneness}
non_null_risks[non_null_risks=='Yes'] <- 1
non_null_risks[non_null_risks=='No'] <- 0

change_type <- c('smoker', 'alcohol', 'gamble', 'skydiving', 'speeding', 'cheated')

non_null_risks <- non_null_risks %>% 
  mutate_at(change_type, as.integer)

non_null_risks$doneness <- factor(non_null_risks$doneness, levels = c('Rare', 'Medium rare', 'Medium', 'Medium Well', 'Well'))

non_null_risks <- non_null_risks %>% 
  mutate(total_risks = smoker + alcohol + gamble + skydiving + speeding + cheated)
```

Using a crosstab view, we can determine how people like their steak within each region 

```{r crosstab}
knitr::kable(crosstab(non_null_risks, x = us_region, y = doneness, weight = total_risks))
```



```{r grouped-us_region-doneness}
grouped <- crosstab(non_null_risks, x = us_region, y = doneness, weight = total_risks, format = 'long')

grouped %>% 
  ggplot(aes(x = pct, y = us_region, fill = doneness)) +
  geom_bar(stat = 'identity', position = position_fill(reverse = TRUE)) +
  xlab("Total Region %") +
  ylab(element_blank()) +
  geom_text(aes(label = paste0(round(pct, 0), '%')), position = position_fill(reverse = TRUE, vjust = 0.5), size = 2.5, color = 'black', fontface = 'bold') +
  scale_x_continuous(labels = scales::percent) +
 theme(legend.position = "top", legend.title = element_blank()) +
  labs(title = 'Steak Doneness by US Region') +
  scale_y_discrete(limits=rev) +
  scale_fill_brewer(palette="RdBu")
```

Surprisingly, not every region is represented by people who prefer their steaks rare. 

## Recommendations
Given how the survey was taken, more information would be beneficial to help determine riskiness. Each question only offered the options of Yes/No, however, the degree of risk may vary based on frequency. A person who drinks alcohol, may drink a few glasses per week, but another person may drink several per day. The survey is also at risk of self-selection bias and does not account for random sampling. Lastly, are certain race/religions more prone to how they like to eat steak? 